---
title: "Lab 3 - Data Types | Exploring Datasets"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Spring 2023"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

1.  Discuss and navigate different data types in R
2.  Create, manipulate, and explore datasets
3.  Date objects

## Data Types in R

R treats objects differently based on their characteristics. For more information, please see: <https://www.statmethods.net/input/datatypes.html>.

-   **Vectors** 1 dimensional structure that contains elements of the same type. Everything in a vector has to be the same data type. Boulion (sp?) is just TRUE, FALSE. A type of array.

-   **Matrices** 2 dimensional structure (rows and columns) that contains elements of the same type. Can call on specific values in a certain row and column. A type of array. 3 dimensional matrix can include time (stack of matrices)

-   **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

-   **Lists** Ordered collection of elements that can have different modes (aka mixing data types).

-   **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (\# at beginning of line, ---- at end)

```{r Create various data structures}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector; c = concatenate (aka put things together); if you put a different data type in here, R will categorize it based on the lowest common denominator. SO, if you put a character value in the numeric vector, it all becomes a character
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] # this "[]" means get me the third item from that vector, so if you put vector1[5], you get the fifth item from that vector

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # R starts at first column and ascends values down for each row
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) # by row = TRUE means it will ascends values across the row first
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "quatro", "cinco"), 
                                  c("un", "deux", "trois", "quatre"))) # dimnames means dimension names, so can name the rows and                                                                           columns

matrix1[4, ] # If you want to pull things up remember, ROWS come first; [ROW, COLUMN]. Leave columns blank, then it will pull out everything from the row regardless of column
matrix1[ , 3] # Then this pulls out everything in column 3
matrix1[c(12, 14)] # This is now specifying we combine this vector into a matrix.
matrix1[c(12:14)] # When using a ":" this is a slice, where you get the 12th-14th items (12, 13, 14).
matrix1[2:4, 1:3] # Can get a slice of rows and a slice of columns; how matrices are subset by their index value. If you don't specify dimension names, then R will just report out 1,2,3 column and row names

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # Specifying the dimensions and rows; using matrix() function to make a matrix, and dimnames to                                      call names of the dimension
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1 # creating a list of 4 items, and the "name" refers to a key name, which allows us to recall those items by their name. The ";" allows you to do two R statements in one line
list1[[2]] # Same idea of returning an item from the list, but requires [[]] because that's how this one works
list1[['mynumbers']] # This is the same idea as above. When using ' vs " it's exactly the same, just need to use the same one

# Data Frames ---- This can be thought of as a list of vectors. Every column in df is a vector, and have a series of vectors that are all the same length. That way we can refer to any cell in our df by the row it's in and then the column
d <- c(1, 2, 3, 4) # What type of vector?
e <- c("red", "white", "red", NA) # What type of vector? "NA" is to specify "no value" because the vector still needs to be the                                        same length in all vectors of a data frame
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? -> a logical vector
dataframe1 <- data.frame(d,e,f) # turns the vectors "d", "e", and "f" into a data frame
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # 

dataframe1[1:2,] # 
dataframe1[c("ID","Passed")] # 
dataframe1$ID #the "$" retrieves a single column from your data frame based on the name

```

```{r}
x <- 
```




Question: How do the different types of data appear in the Environment tab?

> Answer: They are split into different types, like data frames and individual values

Question: In the R chunk below, write "dataframe1\$". Press `tab` after you type the dollar sign. What happens?

> Answer:

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r DataExercise-1a}
high_temp <- c(64,71,72,63,51,52,61,64,64,67)
low_temp <- c(42,49,60,46,32,32,41,47,52,54)
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius. Use a function to create the two new vectors from your existing ones in Fahrenheit.

```{r DataExercise-1b}
F_to_C <- function(x) { (x-32)*(5/9) } #the function() MAKES a function that you can then NAME by assigning

C_high_temp <- F_to_C(high_temp)
C_low_temp <-  F_to_C(low_temp)

```

Combine your four vectors into a data frame and add informative column names.

```{r DataExercise-1c}
df_Temps <- data.frame(
  "highsF" = high_temp,
  "lowsF" = low_temp,
 "highsC" = C_high_temp,
  "lowsC" = C_low_temp
) # format is "name you want column", and then vector
```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r DataExercise-1d}
summary(df_Temps)

summary(df_Temps$highsF)
sd(df_Temps$highsF)
```

### Date objects

Remember formatting of dates in R:

%d day as number (0-31) %m month (00-12, can be e.g., 01 or 1) %y 2-digit year %Y 4-digit year %a abbreviated weekday %A unabbreviated weekday %b abbreviated month %B unabbreviated month

```{r Coding with dates}
# Adjust date formatting for today
# Write code for three different date formats. 
# An example is provided to get you started.
# (code must be uncommented)
today <- Sys.Date()
format(today, format = "%B")
#format(today, format = "%Y")
#format(today, format = "%d")
#format(today, format = "")

```

### Package lubridate

Install and load the package lubridate into your R session. Lubridate offers fast and user friendly parsing of date-time data. Create a string for today's data and then convert it to R date object using lubridate.

More info on lubridate [here][<https://cran.r-project.org/web/packages/lubridate/lubridate.pdf>].

```{r DateExercise1}
#install.packages("lubridate")
library(lubridate)

#Ex1
str_today <- "2023-feb-7"
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today)
date_obj_today

#Ex2
str_today2 <- "Feb 7, 2023"
#Since the format is month-day-year we will use function mdy()
date_obj_today <- mdy(str_today2)
date_obj_today

#Ex_3 - on your own...
str_juneteenth <- "19 June 1865"
#Since the format is month-day-year we will use function mdy()
date_juneteenth <- dmy(str_juneteenth)
date_juneteenth
```

```{r DateExercise2}
#century issue
str_past <- "55-feb-3"
date_obj_past <- ymd(str_past)
date_obj_past

ifelse(year(date_obj_past) %% 100 > 23, year(date_obj_past)-100, year(date_obj_past))

#Build a function to fix year that is more general than the one discussed in the lesson
fix.early.dates <- function(d, cutoff) {
       m <- year(d) %% 100  #operator %% is a modular division i.e. integer-divide year(d) by 100 and returns the remainder 
       year(d) <- ifelse(m > cutoff, 1900+m, 2000+m)  #this will update year(d), year() is a function that returns the year for a data object
       return(d)
}

fixed_date_obj_past <- fix.early.dates(date_obj_past,cutoff=23) #cutoff could be the current year to be more general or any other depending on data set 
fixed_date_obj_past
```

```{r centuryfix}
#Fix for century issue
str_past <- "55-feb-3"
#Alternative 1
date_obj_past <- fast_strptime(str_past,"%y-%b-%d",cutoff_2000=23L)
date_obj_past

#Alternative 2
date_obj_past2 <- parse_date_time2(str_past,"ymd",cutoff_2000=23L)
date_obj_past2

#Functions ymd(), mdy(), ydm() do not take argument cutoff_2000
```

In some cases when dates are provided as integers, you may need to provide an origin for your dates. For example, excel date could be given as number of days since an origin date. Origin date can be different. When R looks at dates as integers, its origin is January 1, 1970. Check if that is true on your machine.

```{r dateorigin}
#Check if "1970-01-01" is your origin date.
lubridate::origin

```
